Practical Optional Types for Clojure

### 2015-02-19 formal-model.tex
example rewrite

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
\end{figure}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    &\mbox{Expressions} \\
  \end{altgrammar}
  $$
  \caption{Type Hinting and Non-Reflective Terms}
\end{figure}

@@

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes solved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} models how type hints are propagated.

We implicitly resolve reflection before type checking and remaining reflective calls are type errors.

>>>>

{\TClass}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}\\
            &\alt&  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}\\
    &&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
    \ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \

    \mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Type Hinting and Non-Reflective Terms}
  \label{figure:nonreflectivesyntax}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\TNewStatic}

    {\TFieldStatic}

    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{figure:javatyping}
\end{figure}

@@

core.typed uses tools.analyzer to obtain a rich abstract syntax tree that includes resolved
reflection, however to tell the full story we include a rewrite relation 
in figure~\ref{figure:rewrite} to eliminate reflection when sufficient type hints are provided.
Figure~\ref{figure:hintinfer} models how type hints are propagated.

We now have two languages: an internal language that may contain non-reflective Java calls (figure 
  \ref{figure:nonreflectivesyntax}) 
  and an external language that resembles Clojure, where we must rely on type-hints to resolve
  reflective Java calls (figure~\ref{figure:reflectivesyntax}).

For example, R-FieldElimRefl, which rewrites a reflective field call to be non-reflective,
rewrites the target expression and uses the result to infer a class type hint as the field target.
\fieldtypeliteral{} uses the inferred type hint to lookup if the inferred class has the appropriate
field name and returns the actual target and result type hints which we store in the rewritten rule.

We then extract these type hints in T-FieldStatic for typing checking 
(figure~\ref{figure:javatyping}), which converts type hints
with \javatotcliteral{} to obtain a \emph{non-nilable} type to type check the target expression (which prevents
dereferencing nil) and \javatotcnilliteral{} to get a \emph{nilable} return type (because fields can be nil).

Method and constructor calls are similar, but T-MethodStatic and T-NewStatic disallow passing \nil{} as arguments, 
we assume methods can return \nil{} and we assume constructors cannot.

We rewrite to our internal language before type checking and remaining reflective calls are type errors.
